Clustering time series based on probability distributions across temporal granularities

### 2021-11-15 main.Rmd
remove overlap of point and text in figure 9

<!-- Why instance selection -->
Supervised learning uses a training set of known information to categorize new events through instance selection. Instance selection (@olvera2010review) is a method of rejecting instances that are not helpful for classification. This is analogous to subsampling the population along all dimensions of interest such that the sampled data represents the primary features of the underlying distribution. Instance selection in unsupervised learning has received little attention in the literature, yet it could be a useful tool for evaluating model or method performance. There are several ways to approach prototype selection. Following @Fan2021-bq's idea of picking related examples (neighbors) for each instance (anchor), we can first use any dimensionality reduction techniques like MDS or PCA to project the data into a 2D space. Then pick a few "anchor" customers who are far apart in 2D space and pick a few neighbors for each. Unfortunately, this does not ensure that consumers with significant patterns across all variables are chosen. Tours can reveal variable separation that gets hidden in a single variable display better than static projections. Hence we perform a linked tour with a t-SNE layout using the R package `liminal` (@R-liminal) to identify customers who are more likely to have distinct patterns across the variables studied. (Refer to Supplementary article for further details). Figure \ref{fig:hod-ind-group-png} shows the distribution across `hod` (a), `moy`(b) and `wnwd` (c) for the set of chosen $24$ customers that were chosen. Few of these customers have similar distribution across `moy` and some are similar in their `hod` distribution.

@@

The $24$ prototypes <!--and the data subset consisting of $353$ customers--> are clustered using the methodology described in Section \ref{sec:methodology} and results are reported below. In the following plots, the median is shown by a line, and the shaded region shows the area between the  $25^{th}$ and $75^{th}$ percentiles.<!--All customers with the same color represent the same clustered groups. --> Groups by JS-based distances and wpd-based distances are colored differently as they represent different groupings.
While plotting the clusters, RS-based scaling is applied to each customer, so that no customer dominate the shape of the summarized clusters because of magnitude. The plotting scales are not displayed since we want to emphasize comparable patterns rather than scales. The idea is that a customer in a cluster may have low total energy usage, but their behavior may be quite similar to a customer with high usage with respect to distributional pattern or significance across cyclic granularities.

@@

The $sindex$ is plotted against the number of clusters for clustering based on JS-based (JS-NQT) distances in Figure \ref{fig:opt-cluster-tsne-jsd}. The $sindex$ falls from around $0.56$ when $k=3$ to $0.54$ when $k$ increases. In our study, we use $k=4, 5$. Figure \ref{fig:groups-4and5} depicts the summarised distributions across $4$ and  $5$ clusters in (a) and (b) respectively, and assists in characterizing each cluster. A-2, A-3, and A-4 have profiles that correspond to groups B-2, B-3, and B-5, respectively. B-2 (id: 4-9 in Figure \ref{fig:hod-ind-group-png}) and B-5 (id: 21-24) have a `hod` pattern with a typical morning and evening peak, whereas the other groups have no distinct pattern or high variability throughout the day. A-1 is subdivided into B-1 and B-4, each of which has a distinct shape across 'moy' and 'wnwd'. B-4 (id: 16-20) is distinguished by higher energy consumption in the first few months of the year and greater variability in weekend usage. B-1 (id: 1-3) has higher consumption in the middle of the year (winter months) and a similar weekday-weekend inter-quartile range. When $k = 4$ is used, these two groups merge to form A-1, which has a moy profile of higher usage in both the beginning and middle of the year, which is not representative of the individuals. It may be worthwhile to compare Figures \ref{fig:opt-cluster-tsne-jsd} and \ref{fig:hod-ind-group-png} to see if the summarized distributions across groups accurately characterized the groupings. If it has, then the majority of the members of the group should have a similar profile.

@@

Things become far more complicated when we consider a larger data set with more uncertainty, as they do with any clustering problem. Summarizing distributions across clusters with varied or outlying customers can result in a shape that does not represent the group. Furthermore, combining heterogeneous customers may result in similar-looking final clusters that are not effective for visually differentiating them. It is also worth noting that the Weekend/Weekday behavior in the given case does not characterize any cluster. This, however, will not be true for all of the customers in the data set. If more extensive prototype selection is used, resulting in more comprehensive prototypes in the data set, this method might be used to classify the entire data set into these prototype behaviors. However, the goal of this section was to have a few customers that have significant patterns over one or more cyclic granularities, apply our methodology to cluster them, and demonstrate that the method produces useful clusters.

@@

There are few areas to extend this research. First, larger data sets with more uncertainty complicate matters, as is true for any clustering task. Characterizing clusters with varied or outlying customers can result in a shape that does not represent the group. Moreover, integrating heterogeneous consumers may result in visually identical end clusters, which are potentially not useful. Hence, a way of appropriately scaling it up to many customers such that anomalies are removed before clustering would be useful for bringing forth meaningful, compact and separated clusters. Secondly, we have assumed the time series to be stationary, and hence the distributions are assumed to remain constant for the observation period. In reality, however, it might change. For the smart meter example, the distribution for a customer moving to a different house or changing electrical equipment can change drastically. Our current approach can not detect these dynamic changes. Thirdly, it is possible that for a few customers, data for some categories from the list of considered significant granularities are missing. In our application, we have removed those customers and done the analysis but the metrics used should be able to incorporate those customers in the clustering by handling their missing categories. Finally, $wpd$ is computationally heavy even under parallel computation. Future work can make the computations more efficient so that they are easily scalable to a large number of customers. Moreover, experiments can also be run with non-hierarchy based clustering algorithms to verify if these distances work better with other algorithms.

>>>>

<!-- Why instance selection -->
Supervised learning uses a training set of known information to categorize new events through instance selection. Instance selection (@olvera2010review) is a method of rejecting instances that are not helpful for classification. This is analogous to subsampling the population along all dimensions of interest such that the sampled data represents the primary features of the underlying distribution. Instance selection in unsupervised learning has received little attention in the literature. However, it could be a useful tool for evaluating model or method performance. There are several ways to approach prototype selection. Following @Fan2021-bq's idea of picking related examples (neighbors) for each instance (anchor), we can first use any dimensionality reduction techniques like MDS or PCA to project the data into a 2D space. Then pick a few "anchor" customers far apart in 2D space and pick a few neighbors for each. Unfortunately, this does not ensure that consumers with significant patterns across all variables are chosen. Tours can reveal variable separation that gets hidden in a single variable display better than static projections. Hence, we perform a linked tour with a t-SNE layout using the R package `liminal` (@R-liminal) to identify customers who are more likely to have distinct patterns across the variables studied. (Refer to the supplementary article for further details). Figure \ref{fig:hod-ind-group-png} shows the distribution across `hod`, `moy` and `wnwd` for the set of chosen $24$ customers that were chosen. The customers are split into batches of $12$ in (a) and (b). Each row in (a) and that in (b) represents the profile of one customer across different variables. <!--Few of these customers have similar distribution across `moy` and some are similar in their `hod` distribution. -->

@@

The $24$ prototypes <!--and the data subset consisting of $353$ customers--> are clustered using the methodology described in Section \ref{sec:methodology} and results are reported below. In the following plots, the median is shown by a line, and the shaded region shows the area between the  $25^{th}$ and $75^{th}$ percentiles.<!--All customers with the same color represent the same clustered groups. --> Groups are colored differently whenever they represent different groupings. While plotting the clusters, RS-based scaling is applied to each customer, so that no customer dominates the shape of the summarized clusters because of magnitude. The plotting scales are not displayed since we want to emphasize comparable patterns rather than scales. The idea is that a customer in a cluster may have low total energy usage, but their behavior may be quite similar to a customer with high usage with respect to distributional pattern or significance across cyclic granularities.

@@

The $sindex$ is plotted against the number of clusters for clustering based on JS-based (JS-NQT) distances in Figure \ref{fig:opt-cluster-tsne-jsd}. The $sindex$ falls from around $0.56$ when $k=3$ to $0.54$ when $k$ increases. In our study, we use $k=4, 5$. Figure \ref{fig:groups-4and5} depicts the summarised distributions across $4$ and  $5$ clusters in (a) and (b) respectively, and assists in characterizing each cluster. A-2, Groups A-3, and A-4 have profiles that correspond to B-2, B-3, and B-5, respectively. B-2 (id: 4-9 in Figure \ref{fig:hod-ind-group-png}) and B-5 (id: 21-24) have a `hod` pattern with a typical morning and evening peak, whereas the other groups have no distinct pattern or high variability throughout the day. A-1 is subdivided into B-1 and B-4, each of which has a distinct shape across `moy` and `wnwd`. B-4 (id: 16-20) is distinguished by higher energy consumption in the first few months of the year and greater variability in weekend usage. B-1 (id: 1-3) has higher consumption in the middle of the year (winter months) and a similar weekday-weekend inter-quartile range. When $k = 4$ is used, these two groups merge to form A-1, which has a moy profile of higher usage in both the beginning and middle of the year, which is not representative of the individuals. It may be worthwhile to compare Figures \ref{fig:opt-cluster-tsne-jsd} and \ref{fig:hod-ind-group-png} to see if the summarized distributions across groups accurately characterized the groupings. If it has, then the majority of the members of the group should have a similar profile.

@@

Things become far more complicated when we consider a larger data set with more uncertainty, as they do with any clustering problem. Summarizing distributions across clusters with varied or outlying customers can result in a shape that does not represent the group. Furthermore, combining heterogeneous customers may result in similar-looking final clusters that are not effective for visually differentiating them. It is also worth noting that the Weekend/Weekday behavior in the given case does not characterize most selected customers. This, however, will not be true for all of the customers in the data set. If more extensive prototype selection is used, resulting in more comprehensive prototypes in the data set, this method might be used to classify the entire data set into these prototype behaviors. However, the goal of this section was to have a few customers that have significant patterns over one or more cyclic granularities, apply our methodology to cluster them, and demonstrate that the method produces useful clusters.

@@

There are few areas to extend this research. First, larger data sets with more uncertainty complicate matters, as is true for any clustering task. Characterizing clusters with varied or outlying customers can result in a shape that does not represent the group. Moreover, integrating heterogeneous consumers may result in visually identical end clusters, which are potentially not useful. Hence, a way of appropriately scaling it up to many customers such that anomalies are removed before clustering would be useful for bringing forth meaningful, compact and separated clusters. Secondly, we have assumed the time series to be stationary, and hence the distributions are assumed to remain constant for the observation period. In reality, however, it might change. For the smart meter example, the distribution for a customer moving to a different house or changing electrical equipment can change drastically. Our current approach can not detect these dynamic changes. Thirdly, it is possible that for a few customers, data for some categories from the list of considered significant granularities are missing. In our application, we have removed those customers and done the analysis but the metrics used should be able to incorporate those customers with such structured missingness. Finally, $wpd$ is computationally heavy even under parallel computation. Future work can make the computations more efficient so that they are easily scalable to a large number of customers. Moreover, experiments can also be run with non-hierarchy based clustering algorithms to verify if these distances work better with other algorithms.
